---
title: "Normalization"
author: "Pol Castellano-Escuder"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
    toc_depth: 4
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Normalization}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  dpi = 200,
  fig.align = "center"
)
```

# Load POMA

```{r, warning = FALSE, message = FALSE, comment = FALSE}
library(POMA)
library(Biobase) # to subset example data
library(ggplot2) # to tune POMA default plots
library(patchwork) # to show plots together
library(tidyverse) # it's just tidyverse;)
```

# Load Data

Let's create a smaller `MSnSet` object to better explore normalization effects.      

```{r}
# load example data
data("st000284")

# subset example data
my_features <- t(exprs(st000284))[c(1:8, 120:127) ,]

my_target <- pData(st000284)[c(1:8, 120:127) ,]
my_target <- my_target %>% rownames_to_column("ID")

# create a smaller MSnSet object from example data
example_data <- PomaMSnSetClass(features = my_features, target = my_target)
example_data
```

# Normalization in `POMA`

_We can skip the imputation step because example data doesn't has missing values._

Here we will evaluate ALL normalization methods that POMA offers on the same `MSnSet` object to compare them.  

```{r}
none <- PomaNorm(example_data, method = "none", round = 3)
auto_scaling <- PomaNorm(example_data, method = "auto_scaling", round = 3)
level_scaling <- PomaNorm(example_data, method = "level_scaling", round = 3)
log_scaling <- PomaNorm(example_data, method = "log_scaling", round = 3)
log_transformation <- PomaNorm(example_data, method = "log_transformation", round = 3)
vast_scaling <- PomaNorm(example_data, method = "vast_scaling", round = 3)
log_pareto <- PomaNorm(example_data, method = "log_pareto", round = 3)
```

## Normalization effect on data dimensions

When we check for the dimension of the data after normalization we can see that ALL methods have the same effect on data dimension. `PomaNorm` **only** change the data dimension when the data has **features that only have zeros** and when the data has **features with 0 variance**. Only in these two cases `PomaNorm` will remove features of the data, changing the data dimensions.   

```{r}
dim(Biobase::exprs(none))
dim(Biobase::exprs(auto_scaling))
dim(Biobase::exprs(level_scaling))
dim(Biobase::exprs(log_scaling))
dim(Biobase::exprs(log_transformation))
dim(Biobase::exprs(vast_scaling))
dim(Biobase::exprs(log_pareto))
```

## Normalization effect on samples

```{r, fig.height = 8, fig.width = 10, message = FALSE, comment = FALSE}
p1 <- PomaBoxplots(none, group = "samples", jitter = FALSE) +
  ggtitle("Not Normalized") +
  ylab("") +
  theme(axis.text.x = element_blank())

p2 <- PomaBoxplots(auto_scaling, group = "samples", jitter = FALSE) +
  ggtitle("Auto Scaling") +
  ylab("") +
  theme(legend.position = "none",
        axis.text.x = element_blank())

p3 <- PomaBoxplots(level_scaling, group = "samples", jitter = FALSE) +
  ggtitle("Level Scaling") +
  ylab("") +
  theme(legend.position = "none",
        axis.text.x = element_blank())

p4 <- PomaBoxplots(log_scaling, group = "samples", jitter = FALSE) +
  ggtitle("Log Scaling") +
  ylab("") +
  theme(legend.position = "none",
        axis.text.x = element_blank())

p5 <- PomaBoxplots(log_transformation, group = "samples", jitter = FALSE) +
  ggtitle("Log Transformation") +
  ylab("") +
  theme(legend.position = "none",
        axis.text.x = element_blank())

p6 <- PomaBoxplots(vast_scaling, group = "samples", jitter = FALSE) +
  ggtitle("Vast Scaling") +
  ylab("") +
  theme(legend.position = "none",
        axis.text.x = element_blank())

p7 <- PomaBoxplots(log_pareto, group = "samples", jitter = FALSE) +
  ggtitle("Log Pareto") +
  ylab("") +
  theme(legend.position = "none",
        axis.text.x = element_blank())

p1 / (p2 + p3 + p4) / (p5 + p6 + p7)
```

## Normalization effect on features

```{r, fig.height = 8, fig.width = 10, message = FALSE, comment = FALSE}
p8 <- PomaDensity(none, group = "features") +
  ggtitle("Not Normalized") +
  ylab("") +
  theme(axis.text.x = element_blank())

p9 <- PomaDensity(auto_scaling, group = "features") +
  ggtitle("Auto Scaling") +
  ylab("") +
  theme(legend.position = "none",
        axis.text.x = element_blank())

p10 <- PomaDensity(level_scaling, group = "features") +
  ggtitle("Level Scaling") +
  ylab("") +
  theme(legend.position = "none",
        axis.text.x = element_blank())

p11 <- PomaDensity(log_scaling, group = "features") +
  ggtitle("Log Scaling") +
  ylab("") +
  theme(legend.position = "none",
        axis.text.x = element_blank())

p12 <- PomaDensity(log_transformation, group = "features") +
  ggtitle("Log Transformation") +
  ylab("") +
  theme(legend.position = "none",
        axis.text.x = element_blank())

p13 <- PomaDensity(vast_scaling, group = "features") +
  ggtitle("Vast Scaling") +
  ylab("") +
  theme(legend.position = "none",
        axis.text.x = element_blank())

p14 <- PomaDensity(log_pareto, group = "features") +
  ggtitle("Log Pareto") +
  ylab("") +
  theme(legend.position = "none",
        axis.text.x = element_blank())

p8 / (p9 + p10 + p11) / (p12 + p13 + p14)
```

