---
title: "POMA EDA Example"
author: 
- name: Pol Castellano-Escuder
  affiliation: University of Barcelona, Spain.
  email: polcaes@gmail.com
date: "`r BiocStyle::doc_date()`"
output: 
    BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{POMA EDA Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}
biblio-style: apalike
link-citations: true
---

**Compiled date**: `r Sys.Date()`

**Last edited**: 2020-08-02

**License**: `r packageDescription("POMA")[["License"]]`

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.align = "center",
  comment = ">"
)
```

# Installation

Run the following code to install the Bioconductor version of package.        

```{r, eval = FALSE}
# install.packages("BiocManager")
BiocManager::install("POMA")
```

# Load POMA 

```{r, warning = FALSE, message = FALSE, comment = FALSE}
library(POMA)
```

# Automatic EDA Report 

The following function will return an Exploratory Data Analysis (EDA) PDF report. The input object must be an **MSnSet** object.     

```{r, warning = FALSE, message = FALSE, comment = FALSE, eval = FALSE}
data("st000336")
PomaEDA(st000336)
```

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE}
library(patchwork)
library(knitr)
library(dplyr)
library(tibble)
library(ggplot2)
library(reshape2)
library(Biobase)

e <- t(Biobase::exprs(st000336))
target <- Biobase::pData(st000336) %>% rownames_to_column("ID") %>% rename(Group = 2) %>% select(ID, Group)
```

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE}
imputed <- PomaImpute(st000336, method = "knn")
pre_processed <- PomaNorm(imputed, method = "log_pareto") %>%
  PomaOutliers(coef = 1.5)
```

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE}
# zeros
zeros <- data.frame(number = colSums(e == 0, na.rm = TRUE)) %>%
  rownames_to_column("names") %>%
  filter(number != 0)

all_zero <- zeros %>% 
  filter(number == nrow(e))

# missing values
nas <- data.frame(number = colSums(is.na(e))) %>%
  rownames_to_column("names") %>%
  filter(number != 0)

# zero variance
var_zero <- e %>%
  as.data.frame() %>%
  summarise_all(~ var(., na.rm = TRUE)) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("names") %>%
  filter(V1 == 0)
```

# Know your data

  + Your data have **`r nrow(e)`** samples, **`r ncol(e)`** features and **`r length(table(target$Group))`** groups, that are **`r noquote(paste(shQuote(levels(as.factor(target$Group))), collapse=", "))`**. `r ifelse(ncol(Biobase::pData(st000336)) > 1, paste0("Furthermore, **", ncol(Biobase::pData(st000336)) - 1,"** covariates have been found in your data. These covariates are **",noquote(paste(shQuote(paste0(colnames(Biobase::pData(st000336))[2:ncol(Biobase::pData(st000336))])), collapse=", ")),"**."), "")`          

  + A **`r round((sum(is.na(e))/(nrow(e)*ncol(e)))*100, 2)`%** of values in your data are NAs (missing values). `r ifelse(nrow(nas) >= 1, paste0("Variables that have NA values are **",noquote(paste(shQuote(paste0(nas$names," (",nas$number,")")), collapse=", ")),"**."), "")`

  + A **`r round((sum(zeros$number)/(nrow(e)*ncol(e)))*100, 2)`%** of values in your data are zeros. `r ifelse(nrow(zeros) >= 1, paste0("Variables that have zeros are **",noquote(paste(shQuote(paste0(zeros$names," (",zeros$number,")")), collapse=", ")),"**."), "")`
  
  + Removed from the exploratory data analysis **`r nrow(all_zero)`** features that only have zeros. `r ifelse(nrow(all_zero) >= 1, paste0("These variables are **",noquote(paste(shQuote(all_zero$names), collapse=", ")),"**."), "")` 
  
  + Removed from the exploratory data analysis **`r nrow(var_zero)`** features that have zero variance. `r ifelse(nrow(var_zero) >= 1, paste0("These variables are **",noquote(paste(shQuote(var_zero$names), collapse=", ")),"**."), "")`

## Summary Table

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE}
summary_table1 <- data.frame(Samples = nrow(e),
                             Features = ncol(e),
                             Covariates = ncol(Biobase::pData(st000336)) - 1)
summary_table2 <- data.frame(Counts_Zero = sum(zeros$number),
                             Percent_Zero = paste(round((sum(zeros$number)/(nrow(e)*ncol(e)))*100, 2), "%"))
summary_table3 <- data.frame(Counts_NA = sum(is.na(e)),
                             Percent_NA = paste(round((sum(is.na(e))/(nrow(e)*ncol(e)))*100, 2), "%")) 
knitr::kable(summary_table1)
knitr::kable(summary_table2)
knitr::kable(summary_table3)
```

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE, dpi = 300}
if (nrow(nas) >= 1){
  ggplot(nas, aes(reorder(names, number), number, fill = number)) +
    geom_col() +
    ylab("Missing values") +
    xlab("") +
    ggtitle("Missing Value Plot") +
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none")
}
```

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE, dpi = 300}
if (nrow(zeros) >= 1){
  ggplot(zeros, aes(reorder(names, number), number, fill = number)) +
    geom_col() +
    ylab("Zeros") +
    xlab("") +
    ggtitle("Zeros Plot") +
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none")
}
```

## Samples by Group

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE, dpi = 300}
counts <- data.frame(table(target$Group))
colnames(counts) <- c("Group", "Counts")

ggplot(counts, aes(reorder(Group, Counts), Counts, fill = Group)) +
  geom_col() +
  ylab("Counts") +
  xlab("") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

# Normalization Plots

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE, dpi = 300}
indNum <- nrow(Biobase::pData(pre_processed))
jttr <- ifelse(indNum <= 10, TRUE, FALSE)

p1 <- PomaBoxplots(imputed, jitter = jttr) +
  xlab("Samples") +
  ylab("Value") +
  ggtitle("Not Normalized") +
  theme(legend.position = "bottom")

p2 <- PomaBoxplots(pre_processed, jitter = jttr) +
  xlab("Samples") +
  ylab("Value") +
  ggtitle("Normalized ('log_pareto')") +
  theme(legend.position = "bottom")

p1 + p2
```
  
# Group Distribution Plots

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE, dpi = 300}
p3 <- PomaDensity(imputed) +
  ggtitle("Not Normalized")

p4 <- PomaDensity(pre_processed) +
    ggtitle("Normalized ('log_pareto')")

p3/p4
```

# Outlier Detection

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE}
outliers <- st000336 %>% 
  PomaImpute(method = "knn") %>%
  PomaNorm(method = "log_pareto") %>%
  PomaOutliers(do = "analyze", coef = 1.5)
outliers$polygon_plot
```

**`r nrow(outliers$outliers)`** possible outliers detected in your data. `r ifelse(nrow(outliers$outliers) >= 1, paste0("These outliers are **",noquote(paste(shQuote(paste0(outliers$outliers$sample)), collapse=", ")),"**."), "")`

# High Correlated Features (r > 0.97)

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE}
correlations <- PomaCorr(pre_processed)
high_correlations <- correlations$correlations %>% filter(abs(corr) > 0.97)
```

There are **`r nrow(high_correlations)`** high correlated feature pairs in your data. `r ifelse(nrow(high_correlations) >= 1, paste0("These features are **",noquote(paste(shQuote(paste0(high_correlations$Var1, " - " , high_correlations$Var2)), collapse=", ")),"**."), "")`

# Heatmap and Clustering

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE, dpi = 300}
PomaHeatmap(pre_processed)
```

# Principal Component Analysis

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE, dpi = 300}
PomaMultivariate(pre_processed, method = "pca")$scoresplot
```

