---
title: "Exploratory Data Analysis Report"
subtitle: "Generated with POMA `r paste0('(', packageVersion('POMA'), ')')`"
author: '`r paste0(username, institution)`'
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    number_sections: true
---

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE}
# This file is part of POMA.

# POMA is free software: you can redistribute it and/or modify
# it under the terms of the GNU General Public License as published by
# the Free Software Foundation, either version 3 of the License, or
# (at your option) any later version.

# POMA is distributed in the hope that it will be useful,
# but WITHOUT ANY WARRANTY; without even the implied warranty of
# MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the
# GNU General Public License for more details.

# You should have received a copy of the GNU General Public License
# along with POMA. If not, see <https://www.gnu.org/licenses/>.

library(POMA)

e <- t(SummarizedExperiment::assay(data))
target <- SummarizedExperiment::colData(data) %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column("ID") %>% 
  dplyr::rename(Group = 2) %>% 
  dplyr::select(ID, Group)
```

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE}
if(clean_outliers){
  imputed <- PomaImpute(data, method = imputation)
  pre_processed <- PomaNorm(imputed, method = normalization) %>%
    PomaOutliers(coef = coeff_outliers)
} else {
  imputed <- PomaImpute(data, method = imputation)
  pre_processed <- PomaNorm(imputed, method = normalization)
}
```

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE}
# zeros
zeros <- data.frame(number = colSums(e == 0, na.rm = TRUE)) %>%
  tibble::rownames_to_column("names") %>%
  dplyr::filter(number != 0)

all_zero <- zeros %>% 
  dplyr::filter(number == nrow(e))

# missing values
nas <- data.frame(number = colSums(is.na(e))) %>%
  tibble::rownames_to_column("names") %>%
  dplyr::filter(number != 0)

# zero variance
var_zero <- e %>%
  as.data.frame() %>%
  dplyr::summarise_all(~ var(., na.rm = TRUE)) %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column("names") %>%
  dplyr::filter(V1 == 0)
```

# Know your data

  + Your data have **`r nrow(e)`** samples, **`r ncol(e)`** features and **`r length(table(target$Group))`** groups, that are **`r noquote(paste(shQuote(levels(as.factor(target$Group))), collapse=", "))`**. `r ifelse(ncol(SummarizedExperiment::colData(data)) > 1, paste0("Furthermore, **", ncol(SummarizedExperiment::colData(data)) - 1,"** covariates have been found in your data. These covariates are **",noquote(paste(shQuote(paste0(colnames(SummarizedExperiment::colData(data))[2:ncol(SummarizedExperiment::colData(data))])), collapse=", ")),"**."), "")`          

  + A **`r round((sum(is.na(e))/(nrow(e)*ncol(e)))*100, 2)`%** of values in your data are NAs (missing values). `r ifelse(nrow(nas) >= 1, paste0("Variables that have NA values are **",noquote(paste(shQuote(paste0(nas$names," (",nas$number,")")), collapse=", ")),"**."), "")`

  + A **`r round((sum(zeros$number)/(nrow(e)*ncol(e)))*100, 2)`%** of values in your data are zeros. `r ifelse(nrow(zeros) >= 1, paste0("Variables that have zeros are **",noquote(paste(shQuote(paste0(zeros$names," (",zeros$number,")")), collapse=", ")),"**."), "")`
  
  + Removed from the exploratory data analysis **`r nrow(all_zero)`** features that only have zeros. `r ifelse(nrow(all_zero) >= 1, paste0("These variables are **",noquote(paste(shQuote(all_zero$names), collapse=", ")),"**."), "")` 
  
  + Removed from the exploratory data analysis **`r nrow(var_zero)`** features that have zero variance. `r ifelse(nrow(var_zero) >= 1, paste0("These variables are **",noquote(paste(shQuote(var_zero$names), collapse=", ")),"**."), "")`

## Summary Tables

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE}
summary_table1 <- data.frame(Samples = nrow(e),
                             Features = ncol(e),
                             Covariates = ncol(SummarizedExperiment::colData(data)) - 1)

summary_table2 <- data.frame(Number_Zeros = sum(zeros$number),
                             Percentage_Zeros = paste(round((sum(zeros$number)/(nrow(e)*ncol(e)))*100, 2), "%"))

summary_table3 <- data.frame(Number_Missings = sum(is.na(e)),
                             Percentage_Missings = paste(round((sum(is.na(e))/(nrow(e)*ncol(e)))*100, 2), "%"))

summary_table1 
summary_table2
summary_table3
```

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE, dpi = 300}
if (nrow(nas) >= 1){
  ggplot2::ggplot(nas, ggplot2::aes(reorder(names, number), number, fill = number)) +
    ggplot2::geom_col() +
    ggplot2::labs(x = NULL,
                  y = "Missing values",
                  title = "Missing Value Plot") +
    ggplot2::theme_bw() + 
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
                   legend.position = "none") +
    ggplot2::scale_fill_viridis_c(begin = 0, end = 0.8)
}
```

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE, dpi = 300}
if (nrow(zeros) >= 1){
  ggplot2::ggplot(zeros, ggplot2::aes(reorder(names, number), number, fill = number)) +
    ggplot2::geom_col() +
    ggplot2::labs(x = NULL,
              y = "Zeros",
              title = "Zeros Plot") +
    ggplot2::theme_bw() + 
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
                   legend.position = "none") +
    ggplot2::scale_fill_viridis_c(begin = 0, end = 0.8)
}
```

## Samples by Group

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE, dpi = 300}
counts <- data.frame(table(target$Group))
colnames(counts) <- c("Group", "Counts")

ggplot2::ggplot(counts, ggplot2::aes(reorder(Group, Counts), Counts, fill = Group)) +
  ggplot2::geom_col() +
  ggplot2::labs(x = NULL,
                y = "Counts") +
  ggplot2::theme_bw() + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
                 legend.position = "none") +
  ggplot2::scale_fill_viridis_d(begin = 0, end = 0.8)
```

# Normalization Plots

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE, dpi = 300}
indNum <- nrow(SummarizedExperiment::colData(pre_processed))
jttr <- ifelse(indNum <= 10, TRUE, FALSE)

p1 <- PomaBoxplots(imputed, 
                   jitter = jttr, 
                   label_size = 8,
                   legend_position = "bottom") +
  ggplot2::labs(x = "Samples",
                y = "Value",
                title = "Not Normalized") 

p2 <- PomaBoxplots(pre_processed, 
                   jitter = jttr, 
                   label_size = 8,
                   legend_position = "bottom") +
  ggplot2::labs(x = "Samples",
                y = "Value",
                title = paste0("Normalized (", normalization, ")"))

p1
p2
```

# Group Distribution Plots

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE, dpi = 300}
p3 <- PomaDensity(imputed) +
  ggplot2::ggtitle("Not Normalized")

p4 <- PomaDensity(pre_processed) +
  ggplot2::ggtitle(paste0("Normalized (", normalization, ")"))

p3
p4
```

# Outlier Detection

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE}
outliers <- data %>% 
  PomaImpute(method = imputation) %>%
  PomaNorm(method = normalization) %>%
  PomaOutliers(do = "analyze", coef = coeff_outliers)
outliers$polygon_plot
```

**`r nrow(outliers$outliers)`** possible outliers detected in your data. `r ifelse(nrow(outliers$outliers) >= 1, paste0("These outliers are **",noquote(paste(shQuote(paste0(outliers$outliers$sample)), collapse=", ")),"**."), "")`

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE}
if(nrow(outliers$outliers) >= 1){
  outliers$outliers
  }
```

# High Correlated Features (r > 0.97)

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE}
correlations <- PomaCorr(pre_processed, 
                         label_size = 8)

high_correlations <- correlations$correlations %>% 
  dplyr::filter(abs(corr) > 0.97)
```

There are **`r nrow(high_correlations)`** high correlated feature pairs in your data. `r ifelse(nrow(high_correlations) >= 1, paste0("These features are **",noquote(paste(shQuote(paste0(high_correlations$feature1, " - " , high_correlations$feature2)), collapse=", ")),"**."), "")`

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE, dpi = 300, fig.align = 'center'}
correlations$corrplot
```

# Heatmap and Clustering

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE, dpi = 300}
PomaHeatmap(pre_processed, sample_names = FALSE)
```

# Principal Component Analysis

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE, dpi = 300}
PomaMultivariate(pre_processed, method = "pca", ellipse = FALSE)$scoresplot
```

# Uniform Manifold Approximation and Projection Clustering

```{r, echo = FALSE, warning = FALSE, comment = NA, message = FALSE, dpi = 300}
PomaUMAP(pre_processed, 
         hdbscan_minpts = 5,
         show_clusters = TRUE)$umap_plot
```

